---
title: "MFA Application on Music Data"
output: html_notebook
---
### Part 1: Using MFA

Setting up the data

```{r}
features <- read.csv('features_30_sec.csv')
num_features <- features[,!(names(features) %in% c("label", "filename"))]

df <- num_features

```



Applying the MFA model on the raw data, using scaling without centering.
The MFA model requires the following input:
1. Range of values of Q
2. Range of values of G

Here I have use 6-12 clusters as the exploratory analysis suggested this.
I have used 2:15 factors as the number of features in the reduced dimension space are around this number.
```{r}

set.seed(123)
sim_IMFA_raw <- mcmc_IMIFA(df,
                       method = 'MFA',
                       centering= FALSE, 
                       scaling= "unit",
                       n.iters = 100,
                       storage = storeControl(score.switch = TRUE),
                       range.Q = 2:15,
                       range.G = 6:12,
                       verbose=TRUE)
```


Viewing the results of the MFA clustering.

```{r}
set.seed(123)
true_labels <- features$label
res_MFA <- get_IMIFA_results(sim_IMFA_raw)
summary(res_MFA)
```

The results suggest the following:

1. 6 clusters are formed from the GTZAN data set.
2. Each clusters used 2 factors to represent the songs inside it.

Table of results: MFA clusters vs Original labels

```{r}
table(res_MFA$Clust$MAP, true_labels)
```


```{r}
library(ggplot2)

# Assuming you have the necessary libraries loaded and the data prepared

# Create a data frame
plot_data <- data.frame(
  FA1 = res_MFA$Scores$post.eta[, 1],
  FA2 = res_MFA$Scores$post.eta[, 2],
  Cluster = factor(res_MFA$Clust$MAP)
)

# Scatter plot of PC1 vs PC2 with ggplot2
ggplot(plot_data, aes(x = FA1, y = FA2, color = Cluster)) +
  geom_point(pch = 16) +
  labs(title = "MFA Raw: FA1 vs FA2")
```
Now looking at the factor loadings in each cluster specifically to understand which features define each cluster

```{r}

pheatmap(res_MFA$Loadings$post.load$Cluster1, 
         cluster_rows = F, cluster_cols = F,
         main = paste0("MFA Cluster 1 Factor Loadings"))

pheatmap(res_MFA$Loadings$post.load$Cluster2, 
         cluster_rows = F, cluster_cols = F, 
         main = paste0("MFA Cluster 2 Factor Loadings"))

pheatmap(res_MFA$Loadings$post.load$Cluster3, 
         cluster_rows = F, cluster_cols = F, 
         main = paste0("MFA Cluster 3 Factor Loadings"))

pheatmap(res_MFA$Loadings$post.load$Cluster4,
         cluster_rows = F, cluster_cols = F,
         main = paste0("MFA Cluster 4 Factor Loadings"))

pheatmap(res_MFA$Loadings$post.load$Cluster5, 
         cluster_rows = F, cluster_cols = F,
         main = paste0("MFA Cluster 5 Factor Loadings"))

pheatmap(res_MFA$Loadings$post.load$Cluster6, 
         cluster_rows = F, cluster_cols = F,
         main = paste0("MFA Cluster 6 Factor Loadings"))
```

The estimated parameters of the MFA clustering process:

```{r}

res_MFA$GQ.results$G

res_MFA$GQ.results$Q


res_MFA$GQ.results$Criteria$BICMs

```
The mixing proportions of each cluster:

```{r}
res_MFA$Clust$post.ratio
```
Here majority of the songs are clustered into cluster 6

The least are seen in cluster 3.

Now viewing the mean and variance of each cluster distribution:

```{r}
pheatmap(res_MFA$Means$mus$Cluster1, cluster_rows = F, cluster_cols = F, main = "Cluster 1")
pheatmap(res_MFA$Means$mus$Cluster2, cluster_rows = F, cluster_cols = F, main = "Cluster 2")
pheatmap(res_MFA$Means$mus$Cluster3, cluster_rows = F, cluster_cols = F, main = "Cluster 3")
pheatmap(res_MFA$Means$mus$Cluster4, cluster_rows = F, cluster_cols = F, main = "Cluster 4")
pheatmap(res_MFA$Means$mus$Cluster5, cluster_rows = F, cluster_cols = F, main = "Cluster 5")
pheatmap(res_MFA$Means$mus$Cluster6, cluster_rows = F, cluster_cols = F, main = "Cluster 6")
```
This gives a clearer perception of how the clusters vary in terms of their features. 



Now visualizing how the clusters are formed among just the two factors overall

```{r}

# Create a data frame combining the principal components and cluster labels
data_for_plot <- data.frame(
  FA1 = res_MFA$Scores$post.eta[,1],
  FA2 = res_MFA$Scores$post.eta[,2],
  Cluster = factor(res_MFA$Clust$MAP)
)


# Create a scatterplot matrix
pairs(data_for_plot[,c("FA1", "FA2")],
      col = data_for_plot$Cluster,
      pch = 19,
      main = "Pairs plot of Factors")

```
Exploring the uncertainty in clustering:

All the songs classified with probability of belonging to that cluster = 1

```{r}
clusters_df <-data.frame(res_MFA$Clust$post.prob)

table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster1==1,])),]$label)

```
```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster2==1,])),]$label)
```
```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster3==1,])),]$label)
```

```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster4==1,])),]$label)
```

```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster5==1,])),]$label)
```

```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster6==1,])),]$label)
```

Now looking at those songs clustered with the lowest probability of belonging to the cluster it was allocated to.

```{r}

which(res_MFA$Clust$uncertainty$v>0.4)
res_MFA$Clust$uncertainty$i

df[which(res_MFA$Clust$uncertainty$v>0.4),]

uncertain_indices <- as.numeric(rownames(df[which(res_MFA$Clust$uncertainty$v>0.4),]))
table(res_MFA$Clust$MAP, true_labels)
table(res_MFA$Clust$MAP[uncertain_indices])
table(res_MFA$Clust$MAP)
table(res_MFA$Clust$MAP[uncertain_indices])/table(res_MFA$Clust$MAP)


```
From this it is clear that most of the songs that see a high level of uncertainty in clustering are all songs that belong to the blues genre,and a few from the classical genre.



## Part 2: Using MIFA

Applying the MIFA function requires the following:
1. Range of values of G


The range given here is the same as in MFA (6 to 12)

```{r}
set.seed(123)
sim_MIFA <- mcmc_IMIFA(df,
                       method = 'MIFA',
                       centering= FALSE, 
                       scaling= "unit",
                       n.iters = 100,
                       storage = storeControl(score.switch = TRUE),
                       range.G = 6:12,
                       verbose=TRUE)

```

Viewing the results of MIFA clustering

```{r}

res_MIFA <- get_IMIFA_results(sim_MIFA)
summary(res_MIFA)

```

```{r}
table(res_MIFA$Clust$MAP, true_labels)

```

Now here each cluster is formed using a different number of factors, that is visualizing each cluster using only the first two factors may not give an accurate understanding of their positions in space. 

```{r}
pheatmap(res_MIFA$Loadings$post.load$Cluster1, 
         cluster_rows = F, cluster_cols = F,
         main = paste0("MIFA Cluster 1 Factor Loadings"))

pheatmap(res_MIFA$Loadings$post.load$Cluster2, 
         cluster_rows = F, cluster_cols = F, 
         main = paste0("MIFA Cluster 2 Factor Loadings"))

pheatmap(res_MIFA$Loadings$post.load$Cluster3, 
         cluster_rows = F, cluster_cols = F, 
         main = paste0("MIFA Cluster 3 Factor Loadings"))

pheatmap(res_MIFA$Loadings$post.load$Cluster4,
         cluster_rows = F, cluster_cols = F,
         main = paste0("MIFA Cluster 4 Factor Loadings"))

pheatmap(res_MIFA$Loadings$post.load$Cluster5, 
         cluster_rows = F, cluster_cols = F,
         main = paste0("MIFA Cluster 5 Factor Loadings"))

pheatmap(res_MIFA$Loadings$post.load$Cluster6, 
         cluster_rows = F, cluster_cols = F,
         main = paste0("MIFA Cluster 6 Factor Loadings"))
```
The prbability of belonging to each cluster (pi)

```{r}
res_MIFA$Clust$post.pi
```
Highest is of cluster 6 and lowest of cluster 3. This is in agreement of the MFA results. 


```{r}
plot(res_MIFA, plot.meth = 'zlabels', zlabels = factor(features$label))
```


```{r}
# Create a data frame combining the principal components and cluster labels
data_for_plot <- data.frame(
  FA1 = res_MIFA$Scores$post.eta[,1],
  FA2 = res_MIFA$Scores$post.eta[,2],
  FA3 = res_MIFA$Scores$post.eta[,3],
  FA4 = res_MIFA$Scores$post.eta[,4],
  Cluster = factor(res_MFA$Clust$MAP)
)


# Create a scatterplot matrix
pairs(data_for_plot[,c("FA1", "FA2","FA3","FA4")],
      col = data_for_plot$Cluster,
      pch = 19,
      main = "Pairs plot of Factors")
```
Exploring the uncertainty in clustering:

All the songs classified with probability of belonging to that cluster = 1

```{r}
clusters_df <-data.frame(res_MIFA$Clust$post.prob)

table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster1==1,])),]$label)

```
```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster2==1,])),]$label)
```
```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster3==1,])),]$label)
```

```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster4==1,])),]$label)
```

```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster5==1,])),]$label)
```

```{r}
table(features[as.numeric(rownames(clusters_df[clusters_df$Cluster6==1,])),]$label)
```

Now exploring those with highest uncertainty

```{r}
which(res_MIFA$Clust$uncertainty$v>0.4)
res_MFA$Clust$uncertainty$i

table(features[which(res_MIFA$Clust$uncertainty$v>0.4),]$label)



```
MIFA says that the genres from which songs see the higest uncertainty belong to blues, country and classical. 
This aligns with MFA, however MIFA is also able to factor in country 

