---
title: "Exploring the Dataset"
output: html_notebook
---

### The data - taking a look

```{r}

features <- read.csv('features_30_sec.csv')

#Removing the non-numeric features
num_features <- features[,!(names(features) %in% c("label", "filename"))]

df <- num_features
```


```{r}
# Convert the "labels" column from list to individual elements
genre_list <- unlist(features$label)

# Calculate the count of each genre
genre_counts <- table(genre_list)

# Create a pastel color palette with at least 10 colors
pastel_colors <- c("#FFB6C1", "#FFD700", "#87CEEB", "#98FB98", "#FFA07A", "#9370DB", "#00FF7F", "#FFC0CB", "#00BFFF", "#FFDAB9")

# If you need more than 10 colors, you can extend the vector with additional pastel color codes.

# Create the pie chart with pastel colors
pie(genre_counts, main = "Genre Distribution", col = pastel_colors)

# Add labels to the pie chart
label_percent <- paste0(round(100 * genre_counts / sum(genre_counts), 1), "%")
label_position <- 2 * cumsum(genre_counts) - 0.5 * genre_counts
text(label_position, labels = paste(names(genre_counts), "\n", label_percent), col = "black")

```


Taking a look at the numerical data


```{r}
summary(df)

```
From here it is clear that scaling might be necessary since there are some features in the scale of 10000 and some in the scale of 0.00x

```{r}

#Scaling without centering
scaled_df <- scale(df, center = FALSE)

summary(scaled_df)

```
3. Visualizing the non-cepstral features.

```{r}

boxplot(df[,1:18], main = "Boxplot of Raw Features")

boxplot(scale(df, center = FALSE)[,1:18], main = "Box Plot of Scaled Features")
```


key takeaway: the data here when scaled without centering shows each feature with a different center,
which means it has more variation to offer, which we can leverage in the model based clustering.
so i have chosen to scale without centering.

4. Visualizing the cepstral features.
```{r}
library(pheatmap)
mfcc_df <- df[,19:58]

pheatmap(scale(mfcc_df,center=FALSE),
         cluster_rows = FALSE, cluster_cols = FALSE,
         main = "MFCC Heatmap")

```


## Exploring the features:

1. Continuos Time signal vs Discrete Time signal
 

```{r}
# Install and load necessary libraries
#install.packages("tuneR")
#install.packages("ggplot2")
library(tuneR)
library(ggplot2)

# Specify the path to your .wav file
wav_file_path <- "C:\\Users\\manya\\Documents\\MIR Thesis\\Plots\\Data\\genres_original\\blues\\blues.00000.wav"

# Read the .wav file
audio <- readWave(wav_file_path)

# Extract the audio data
audio_data <- audio@left

# Create a time vector based on the sampling rate
time <- seq(0, (length(audio_data) - 1) / audio@samp.rate, by = 1 / audio@samp.rate)

# Create a data frame for plotting
df <- data.frame(time = time, amplitude = audio_data)

# Filter data for the first 4 seconds
df <- df[df$time <= 4, ]

# Plot the time domain signal using ggplot2
ggplot(df, aes(x = time, y = amplitude)) +
  geom_line() +
  labs(x = "Time (seconds)", y = "Amplitude") +
  ggtitle("Time Domain Signal of WAV File")

```

```{r}

# Define the desired sampling rate (e.g., GTZAN's sampling rate)
desired_sampling_rate <- 44100

# Calculate the resampling ratio
resampling_ratio <- desired_sampling_rate / audio@samp.rate

# Resample the audio data using linear interpolation
resampled_audio_data <- approx(seq_along(audio_data), audio_data, 
                                seq(1, length(audio_data), by = resampling_ratio))$y

# Create a time vector based on the new sampling rate
time <- seq(0, (length(resampled_audio_data) - 1) / desired_sampling_rate, 
            by = 1 / desired_sampling_rate)

# Create a data frame for plotting
df <- data.frame(time = time, amplitude = resampled_audio_data)

# Filter data for the first 4 seconds
df <- df[df$time <= 4, ]

# Plot the discretized time domain signal for the first 4 seconds using ggplot2
ggplot(df, aes(x = time, y = amplitude)) +
  geom_line() +
  labs(x = "Time (seconds)", y = "Amplitude") +
  ggtitle("Discretized Time Domain Signal (First 4 Seconds) with GTZAN's Sampling Rate")
```
The sampled version looks nearly identical to the time domain signal, the information loss is negligible at this sampling rate.


2. Visualizing the spectrogram

```{r}
library(tuneR)
library(seewave)
audio_2 <- readWave("C:\\Users\\manya\\Documents\\MIR Thesis\\Plots\\Data\\genres_original\\hiphop\\hiphop.00000.wav")
audio_2 <- audio_2@left - mean(audio_2@left)
audio_data <- audio@left-mean(audio@left)

spec <- signal::specgram(x = audio_data, n = 1024, Fs = audio@samp.rate, overlap = 1024 * 0.75)

# normalize and rescale to dB
P <- abs(spec$S)
P <- P/max(P)

out <- pmax(1e-6, P)
dim(out) <- dim(P)
out <- log10(out) / log10(1e-6)

# plot spectrogram
image(x = spec$t, y = spec$f, z = t(out), ylab = 'Freq [Hz]', xlab = 'Time [s]', useRaster=TRUE)
```

3. Calculating the RMS energy

```{r}
rms_energy <- sqrt(mean(audio@left^2))  # Calculate RMS for the left channel

# Print the RMS energy
cat("RMS Energy:", rms_energy, "\n")
```
4. Visualizing feature distributions

```{r}



library(dplyr)
library(ggplot2)

df <- features

# Group observations by label and calculate the mean rms_mean value for each group
grouped_data <- df %>%
  group_by(label) %>%
  summarise(mean_rms_mean = mean(rms_mean))
  arrange(mean_rms_mean)  # Arrange in ascending order



# Create a bar plot to visualize the mean rms_mean for each label
ggplot(grouped_data, aes(x = reorder(label, mean_rms_mean), y = mean_rms_mean, fill = label)) +
  geom_bar(stat = "identity") +
  labs(x = "Genres", y = "Mean RMS Energy", title = "Mean RMS Energy Value by Genres") +
  theme_minimal()
```



```{r}
grouped_data <- df %>%
  group_by(label) %>%
  summarise(mean_harmony = mean(spectral_bandwidth_mean)) %>%
  arrange(mean_harmony)  # Arrange in ascending order

# Create a bar plot to visualize the mean harmony for each label
ggplot(grouped_data, aes(x = reorder(label, mean_harmony), y = mean_harmony, fill = label)) +
  geom_bar(stat = "identity") +
  labs(x = "Label", y = "Mean harmony value", title = "Mean Harmony Value by Label (Ascending)") +
  theme_minimal()
```


```{r}
# Group observations by label and calculate the mean tempo value for each group
grouped_data <- df %>%
  group_by(label) %>%
  summarise(mean_tempo = mean(tempo)) %>%
  arrange(mean_tempo)  # Arrange in ascending order

# Create a bar plot to visualize the mean tempo for each label
ggplot(grouped_data, aes(x = reorder(label, mean_tempo), y = mean_tempo, fill = label)) +
  geom_bar(stat = "identity") +
  labs(x = "Label", y = "Mean tempo value", title = "Mean Tempo Value by Label (Ascending)") +
  theme_minimal()
```
```{r}
library(ggplot2)
library(dplyr)
library(patchwork)


features <- c("spectral_centroid_mean", "spectral_bandwidth_mean", "rolloff_mean", "chroma_stft_mean", "harmony_mean")

# Create a list to store the plots
plots <- list()

# Loop through each feature, calculate mean by label, and create a bar plot
for (feature in features) {
  grouped_data <- df %>%
    group_by(label) %>%
    summarise(mean_feature = mean(!!sym(feature))) %>%
    arrange(mean_feature)  # Arrange in ascending order
  
  plot <- ggplot(grouped_data, aes(x = reorder(label, mean_feature), y = mean_feature, fill = label)) +
    geom_bar(stat = "identity") +
    labs(x = "Label", y = paste(feature, "value"), title = paste(feature)) +
    theme_minimal()
  
  plots[[feature]] <- plot
}

# Combine plots using patchwork
combined_plots <- plots[[1]]
for (i in 2:length(plots)) {
  combined_plots <- combined_plots + plots[[i]]
}

# Display the combined plots
combined_plots

```


```{r}
library(pheatmap)

pheatmap(scaled_df[,19:58], cluster_rows = F,cluster_cols = F,
         main = "MFCC Heatmap")
```






